\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid

### 2014-09-22 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Added seconds to success figure.
By removing the [t] after figures I was able to better space figures in the paper.
Moved small population figure.
Filled in large bits of computational costs section.

\section{Parameter-less Population Pyramid}
\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
performing optimization which does not require the user to provide any parameters. This is
achieved by combining efficient local search with the model building methods of LTGA using
an iteratively constructed hierarchy of populations.

@@

@@

@@

\begin{figure}[t]
  \begin{center}
  \includegraphicsfit{evals-to-success}
  \end{center}
  \caption{Comparison of the median number of evaluations to reach the global optimum for
           the six different optimization methods with respect
           to problem size.  If the median run did not reach the global optimum no data element
           is shown.  Results given on a log-log scale.}
  \label{fig-evals-to-success}
\end{figure}

@@

\begin{figure}[t]
  \begin{center}
  \includegraphicsfit{evals-to-success-range}
  \end{center}
  \caption{Comparison of the upper and lower quartiles of evaluations required
           to reach the global optimum for P3 and LTGA with respect to problem size.}
  \label{fig-evals-to-success-range}
\end{figure}

\section{Fitness Over Time}
\begin{figure}[t]
  \begin{center}
  \includegraphicsfit{fitness-over-time}
  \end{center}
  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
  \label{fig-fitness-over-time}
\end{figure}

@@

\section{Computational Expenses}
\begin{figure}[t]
  \begin{centering}
    \begin{subfigure}{.5\textwidth}
      \begin{centering}
        \includegraphicsfit{rebuilds}
      \end{centering}
      \caption{Rebuilds}
      \label{fig-rebuilds}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
      \begin{centering}
        \includegraphicsfit{donations}
      \end{centering}
      \caption{Donations}
      \label{fig-donations}
    \end{subfigure}
  \end{centering}
  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
\end{figure}

TODO Include graphic of rebuilds per evaluation. Figure~\ref{fig-rebuilds}.

TODO Include graphic of donations per evaluation. Figure~\ref{fig-donations}.

TODO Include a table of $O(N)$ seconds per evaluation and total seconds for all 4 model building techniques on all 7 problems.

>>>>

\section{Parameter-less Population Pyramid}
\label{sec-p3}
\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
performing optimization which does not require the user to provide any parameters. This is
achieved by combining efficient local search with the model building methods of LTGA using
an iteratively constructed hierarchy of populations.

@@

\begin{figure}
  \begin{center}
  \includegraphicsfit{evals-to-success}
  \end{center}
  \caption{Comparison of the median number of evaluations to reach the global optimum for
           the six different optimization methods with respect
           to problem size.  If the median run did not reach the global optimum no data element
           is shown.  Results given on a log-log scale.}
  \label{fig-evals-to-success}
\end{figure}

@@

\begin{figure}
  \begin{center}
  \includegraphicsfit{evals-to-success-range}
  \end{center}
  \caption{Comparison of the upper and lower quartiles of evaluations required
           to reach the global optimum for P3 and LTGA with respect to problem size.}
  \label{fig-evals-to-success-range}
\end{figure}

@@

@@

\section{Fitness Over Time}
\begin{figure}
  \begin{center}
  \includegraphicsfit{fitness-over-time}
  \end{center}
  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
  \label{fig-fitness-over-time}
\end{figure}

@@

\section{Computational Expenses}
\begin{figure}[t]
  \begin{centering}
    \begin{subfigure}{.5\textwidth}
      \begin{centering}
        \includegraphicsfit{rebuilds}
      \end{centering}
      \caption{Rebuilds}
      \label{fig-rebuilds}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
      \begin{centering}
        \includegraphicsfit{donations}
      \end{centering}
      \caption{Donations}
      \label{fig-donations}
    \end{subfigure}
  \end{centering}
  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
  \label{fig-costs}
\end{figure}

When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
provides some insight into how often these two aspects of the algorithm are utilized.

Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding for search.
In order to calculate this value we recorded how many times search rebuilt the model during each run.
Under the minimal assumption that an evaluation costs $O(N)$ time and that model rebuilds cost $O(N^2)$,
Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically linear.
For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this appears to be the case. For both Trap problems
and HIFF there appears to be slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
bounded to no more than twice the algorithmic cost of an evaluation.

When applying a crossover subset, P3 tries random donors from the population until one is found with at least
one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
important note again is that each donation may range in size from a single bit up to $N-1$, with smaller
donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
in P3, but its unlikely to be very high.

\begin{figure}
  \begin{center}
  \includegraphicsfit{seconds-to-success}
  \end{center}
  \caption{Comparison of the median number of seconds to reach the global optimum for
           the six different optimization methods with respect
           to problem size.  If the median run did not reach the global optimum no data element
           is shown.  Results given on a log-log scale.}
  \label{fig-seconds-to-success}
\end{figure}

To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}. Due to the scale
of our experiments, diverse hardware potentially running multiple processes in parallel were used
in order to collect all runs. As such these timings should be considered very skeptically as a number
of sources may contribute to biased error in the results. Under those caveats, the results do suggest
that any increase in P3's algorithmic complexity is offset by its efficient use of evaluations. On
Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT it still appears to find the global
optimum faster than all comparison techniques. On HIFF LTGA is able to match P3's performance almost exactly,
and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
than LTGA but by no more than a constant factor.

When considering wall clock time hBOA and Parameter-less hBOA perform significantly worse than when using
evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while under reasonable
assumptions P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
As P3 and LTGA require a similar asymptotic complexity per
evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.

These results suggest that even for minimally expensive fitness functions P3 can find the global optimum
at least as quickly in seconds as its optimally tuned competitors. However, due to the potential noise in our data collection
we are hesitant to draw too strong a conclusion.
